\ttitle

### 2013-10-25 Chapters/Chapter3.tex
Draft, sent to Anne

This chapter discusses the concepts and algorithms that will be used as a basis for the proposed method, as discussed in Chapter \ref{Chapter4}.
The first sections formulates the problem of change detection and relates it to outlier and novelty detection.
It transforms the traditional problem to change detection for times series data.
In Section \ref{sec:one_class_classification} the problem of \acrlong{occ} is discussed and shows a number of implementations.
Two \acrlong{svm}-based methods, \gls{nu-svm} and \gls{svdd}, are further disussed in section \ref{sec:one_class_svm}.
\TODO{add \ref{sec:svm_parameters}}
For the application to the problem of \acrlong{har}, the characteristics of the collected sensor data is discussed in \ref{sec:sensor_data_characteristics}.
Finally a theoretic set of quality metrics is discussed in Section \ref{sec:quality_metrics}.

\input{./Chapters/Chapter3/change_detection_time_series}
\input{./Chapters/Chapter3/one_class_classification}
\input{./Chapters/Chapter3/one_class_svm}
\input{./Chapters/Chapter3/svm_parameters}
\input{./Chapters/Chapter3/sensor_data_characteristics}
\input{./Chapters/Chapter3/quality_metrics}

>>>>

This chapter discusses the concepts and algorithms that will be used as a basis for the proposed method, as discussed in Chapter \ref{Chapter4}.
The first sections formulates the problem of change detection and relates it to outlier and novelty detection.
It transforms the traditional problem to change detection for times series data.
In Section \ref{sec:one_class_classification} the problem of \acrlong{occ} is discussed and shows a number of implementations.
Two \acrlong{svm}-based methods, \gls{nu-svm} and \gls{svdd}, are further disussed in section \ref{sec:one_class_svm}.
For the application to the problem of \acrlong{har}, the characteristics of the collected sensor data is discussed in \ref{sec:sensor_data_characteristics}.
Finally a theoretic set of quality metrics is discussed in Section \ref{sec:quality_metrics}.

\input{./Chapters/Chapter3/change_detection_time_series}
\input{./Chapters/Chapter3/one_class_classification}
\input{./Chapters/Chapter3/one_class_svm}
\input{./Chapters/Chapter3/sensor_data_characteristics}
\input{./Chapters/Chapter3/quality_metrics}

### 2013-10-25 Chapters/Chapter3/one_class_svm.tex
Draft, sent to Anne

A \gls{svm} can create a boundary between linear-separable data points, making it a non-probabilistic binary linear classifier.
More flexible non-linear boundaries can be obtained by the use of a non-linear function $\phi(x)$, as illustrated in Figure \ref{fig:kernel_mapping}.
This function maps the input data from space $\mathcal{I}$ to a higher dimensional space $\mathcal{F}$.
The \gls{svm} can create a linear separating hyperplane in the space $\mathcal{F}$ that separates the data points from the two classes.
When the hyperplane is projected to the (lower) original input space $\mathcal{I}$ it creates a non-linear separating curve.
The mapping and projection of data points can be efficient (and implicit) performed by using the kernel trick, which is discussed in section \ref{subsec:kernel_trick}.

@@

@@

This minimization problem can be solved (using \gls{qp}) and transformed to its Lagrange dual formulation.
\\ \TODO{Add explanation why to use dual, which is: because the problem formulation scales with the number of samples $n$ and not the dimesionality $d$ of the samples.
More in Section 9.3 of \cite{cherkassky2007learning}.} \\
To do so, the Lagrange multipliers $a_i >= 0$ are introduced and the decision function becomes:
\begin{equation}\label{eq:svm_lagrange}
  f(x) = \operatorname{sgn}( \sum_{i=1}^n \alpha_i y_i K(x, x_i) + b),
\end{equation}
where $K(x, x_i) = \phi(x)^T\phi(x_i)$ (which is further discussed in section \ref{subsec:kernel_trick}).
Here every data point $\mathcal{I}$ for which $a_i > 0$ is weighted in the decision function and thus ``supports'' the classification machine: hence the name ``\acrlong{svm}''.
Since it is shown that under certain circumstances \glspl{svm} show an equality to sparse representations \cite{girosi1998equivalence,smola1998connection}, there will often be relatively few Lagrange multipliers with a non-zero value.

@@

\subsection{One-Class Support Vector Machines}\label{subsec:one-class-svm}
The classical implementation of a \gls{svm} is to classify a dataset in two distinct classes.
This is a common usecase, although sometimes there is no training data for both classes available.
Still, one would like to classify new data points as regular, or in-class, or out-of-class, \eg in the case of a novelty detection.
With that problem only data from one class is available and the objective is to recognize new data points that are not part of that class.
This problem is closely related to density estimation.
In that context, the problem can be the following.
Given an underlying probabilty distribution $P$ the goal is to find a subset $S$ for which the probability that a data point from $P$ lies outside $S$ equals some predetermined value $\nu$ between $0$ and $1$ \cite{scholkopf1999support}.

@@

\TODO{Add something about robustness, discusses in section 9.2 from \cite{cherkassky2007learning}}

@@

The constructed hypersphere is characterized by a center $\mathbf{a}$ and a radius $R > 0$ as distance from the center to (any data point that is a \gls{sv} on) the boundary, for which the volume, and thus the radius $R$, will be minimized.
The center $\mathbf{a}$ is a linear combination of of the support vectors.
Like the classical \gls{svm} and \gls{svdd} it can be required that all the distances from the data points $x_i$ to the center $\mathbf{a}$ are strict less then $R$ (or equivalent measure, to create a hard margin) or soft margins are allowed by using slack variables $\xi_i$.
In the case of a soft margin, the penalty is determined by $C$ and the minimization is expressed as Equation \ref{eq:svdd_objective}.
This principle is illustrated in the right image or Figure \ref{fig:nu-vs-svdd}.
Instead of a separating hyperplane, constructed by \gls{nu-svm} and illustrated on the left of the Figure, the \gls{svdd} creates a hypersphere (in the illustration a cricle) around the data points.
By using kernel functions (\eg the \gls{rbf}) the hyperspheres in the high dimensional feature space $\mathcal{F}$ corresponds to a flexible and tight enclosing boundary in input space $\mathcal{I}$, illustrated in Figure \ref{fig:svdd-boundary}.

@@

A common class for kernel function is the \acrlong{rbf}.
These function are of the form
\begin{equation}
  g(x) = \operatorname*{sign} \left(  \sum_{i=1}^{n} \alpha_i \operatorname*{exp} \left\{ \frac{|\vectorsym{x} - \vectorsym{x}_i|^2}{\sigma^2} \right\} \right),
\end{equation}
where $\sigma$ defines the width.
The inner product kernel is of the form
\begin{equation}
  H(\vectorsym{x}, \vectorsym{x}') = \operatorname*{exp} \left\{ - \frac{| \vectorsym{x} - \vectorsym{x}'|^2}{\sigma^2} \right\}.
\end{equation}

Note that with the \gls{rbf} class of kernel functions the number of basis functions, the center parameters that correspond to the support vectors, and the weights in the output layer are all automatically determined via the optimal hyperplane \cite{cherkassky2007learning}.
The width parameter is equal for all basis functions and determined a priori.
Other examples of kernel function classes are \emph{pylonomial functions} of degree $q$, \emph{neural networks}, \emph{splines} of order $m$ with $b$ nodes, and \emph{Fourier expansions}.
\TODO{More on basis functions. \cite{cherkassky2007learning}, chapter 9.4}

$k(\mathbf{x}_i, \mathbf{x})$ is a kernel function computing a dot product in feature space, introduced by Aizerman \etal \cite{aizerman1964theoretical}.
``Training a \gls{svm} with Gaussian \gls{rbf} kernels corresponds to minimizing the specific cost function with a regularization operator'' \cite{smola1998connection}. \\
``Gaussian kernels tend to yield good performance under general smoothness assumptions and should be considered especially of no additional knowledge of the data is available.'' \cite{smola1998connection}. \\
``Choosing a small width of the kernels leads to high generalization error as it effectively decouples the separate basis functions of the kernel expansion into very localized functions which is equivalent to memorizing the data, whereas a wide kernel tends to oversmooth.'' \cite{smola1998connection}. \\

@@

@@

``Learning from Data'' \cite{cherkassky2007learning}. 2007, 1547 (book).

>>>>

A \gls{svm} can create a boundary between linear-separable data points, making it a non-probabilistic binary linear classifier.
More flexible non-linear boundaries can be obtained by the use of a non-linear function $\phi(x)$, as illustrated in Figure \ref{fig:kernel_mapping}.
This function maps the input data from space $\mathcal{I}$ to a higher dimensional space $\mathcal{F}$.
The \gls{svm} can create a linear separating hyperplane in the space $\mathcal{F}$ that separates the data points from the two classes.
When the hyperplane is projected to the (lower) original input space $\mathcal{I}$ it creates a non-linear separating curve.
This is illustrated in Figure \ref{fig:svm_mapping_spaces}.
The mapping and projection of data points can be efficient (and implicit) performed by using the kernel trick, which is discussed in section \ref{subsec:kernel_trick}.

@@

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{./Figures/chapter3/svm_mapping_spaces.pdf}
  \caption[Mapping spaces in \gls{svm}]{The input vector $\vectorsym{x}$ from input space $\mathcal{I}$ is mapped by a non-linear function $\phi(\vectorsym{x})$ to a feature vector $\vectorsym{z}$ in the high dimensional feature space $\mathcal{F}$. The weights of $\vectorsym{w}$ create a linear separating hypeplace, which maps the high dimensional vector to the predicted outcome $\vectorsym{y}$. \TODO{Is this figure obsolete? Figure \ref{fig:kernel_mapping} shows almost the same}}
  \label{fig:svm_mapping_spaces}
\end{figure}

@@

This minimization problem can be solved (using \gls{qp}) and transformed to its (Lagrange) dual formulation.
In the dual formulation the problem scales with the number of training examples $n$ instead of the dimensionality $d$ of the samples.
Solving this problem directly in the high dimensional feature space $\mathcal{F}$ makes it untractable.
The linear approximation function corresponds to the kernel function in the dual formulation.
Solving this dual formulation is equivalent to solving the primal formulation \cite{cherkassky2007learning}.
In the dual formulation the Lagrange multipliers $a_i >= 0$ are introduced and the decision function becomes:
\begin{equation}\label{eq:svm_lagrange}
  f(x) = \operatorname{sgn}( \sum_{i=1}^n \alpha_i y_i K(x, x_i) + b),
\end{equation}
where $K(x, x_i) = \phi(x)^T\phi(x_i)$ (which is further discussed in section \ref{subsec:kernel_trick}).
Here every data point $\mathcal{I}$ for which $a_i > 0$ is weighted in the decision function and thus ``supports'' the classification machine: hence the name ``\acrlong{svm}''.
Since it is shown that under certain circumstances \glspl{svm} show an equality to sparse representations \cite{girosi1998equivalence,smola1998connection}, there will often be relatively few Lagrange multipliers with a non-zero value.

@@

\subsection{One-Class Support Vector Machines}\label{subsec:one-class-svm}
The classical implementation of a \gls{svm} is to classify a dataset in two distinct classes.
This is a common usecase, although sometimes there is no training data for both classes available.
Still, one would like to classify new data points as regular, or in-class, or out-of-class, \eg in the case of a novelty detection.
With that problem only data from one class is available and the objective is to recognize new data points that are not part of that class.
This unsupervised learning problem is closely related to density estimation.
In that context, the problem can be the following.
Given an underlying probabilty distribution $P$ the goal is to find a subset $S$ for which the probability that a data point from $P$ lies outside $S$ equals some predetermined value $\nu$ between $0$ and $1$ \cite{scholkopf1999support}.

@@

\TODO{Add something about robustness, discusses in section 9.2 from \cite{cherkassky2007learning,chen2004m}}

@@

The constructed hypersphere is characterized by a center $\mathbf{a}$ and a radius $R > 0$ as distance from the center to (any data point that is a \gls{sv} on) the boundary, for which the volume, and thus the radius $R$, will be minimized.
The center $\mathbf{a}$ is a linear combination of of the support vectors.
Like the classical \gls{svm} and \gls{svdd} it can be required that all the distances from the data points $x_i$ to the center $\mathbf{a}$ are strict less then $R$ (or equivalent measure, to create a hard margin) or soft margins are allowed by using slack variables $\xi_i$.
In the case of a soft margin, the penalty is determined by $C$ and the minimization is expressed as Equation \ref{eq:svdd_objective}.
This principle is illustrated in the right image or Figure \ref{fig:nu-vs-svdd}.
Instead of a separating hyperplane, constructed by \gls{nu-svm} and illustrated on the left of the Figure, the \gls{svdd} creates a hypersphere (in the illustration a cricle) around the data points.
By using kernel functions (\eg the \gls{rbf}) the hyperspheres in the high dimensional feature space $\mathcal{F}$ corresponds to a flexible and tight enclosing boundary in input space $\mathcal{I}$.
Possible resulting closed boundaries are illustrated in Figure \ref{fig:svdd-boundary}.
This enclosing boundary is obtained by minimizing:

@@

A common class for kernel function is the \acrlong{rbf}.
These function are of the form
\begin{equation}
  g(x) = \operatorname*{sign} \left(  \sum_{i=1}^{n} \alpha_i \operatorname*{exp} \left\{ \frac{|\vectorsym{x} - \vectorsym{x}_i|^2}{\sigma^2} \right\} \right),
\end{equation}
where $\sigma$ defines the width.
The inner product kernel is of the form
\begin{equation}
  K(\vectorsym{x}, \vectorsym{x}') = \operatorname*{exp} \left\{ - \frac{| \vectorsym{x} - \vectorsym{x}'|^2}{\sigma^2} \right\}.
\end{equation}

Note that with the \gls{rbf} class of kernel functions the number of basis functions, the center parameters that correspond to the support vectors, and the weights in the output layer are all automatically determined via the optimal hyperplane \cite{cherkassky2007learning}.
The width parameter $\sigma$ is equal for all basis functions and is set a priori and determines the flexibility and complexity of the boundary.
In Section \ref{subsec:svm_model_parameters} this (hyper)parameter for a \gls{svm} is further discussed.
Other examples of kernel function classes are \emph{pylonomial functions} of degree $q$, \emph{neural networks}, \emph{splines} of order $m$ with $b$ nodes, and \emph{Fourier expansions}.
\TODO{More on basis functions. \cite{cherkassky2007learning}, chapter 9.4}

-- Quotes and refs to use --
$k(\mathbf{x}_i, \mathbf{x})$ is a kernel function computing a dot product in feature space, introduced by Aizerman \etal \cite{aizerman1964theoretical}.
``Training a \gls{svm} with Gaussian \gls{rbf} kernels corresponds to minimizing the specific cost function with a regularization operator'' \cite{smola1998connection}. \\
``Gaussian kernels tend to yield good performance under general smoothness assumptions and should be considered especially of no additional knowledge of the data is available.'' \cite{smola1998connection}. \\
``Choosing a small width of the kernels leads to high generalization error as it effectively decouples the separate basis functions of the kernel expansion into very localized functions which is equivalent to memorizing the data, whereas a wide kernel tends to oversmooth.'' \cite{smola1998connection}. \\

@@

\subsection{SVM model parameters}\label{subsec:svm_model_parameters}
\gls{svm}-model selectiong and tuning depends on two type of parameters \cite{cherkassky2007learning}:
\begin{enumerate}
  \item Parameters controlling the `margin' size,
  \item Model parameterization, \eg the kernel type and complexity parameters.
  For the \gls{rbf} kernel the width parameter determines the model complexity.
\end{enumerate}

In case of a \gls{rbf} kernel, the width parameter $\sigma$ determines the flexibility and complexity of the boundary.
The value of this parameter greatly determines the outcomes of the algorithm (\eg \gls{svdd}) as illustrated in Figure \ref{fig:svdd-boundary}.
With a small value for the kernel width $\sigma$, each data point will tend to be used as a support vector and will represent a small Gaussian.
The solution then corresponds with standard non-parametric density estimation.
With a large value for the width $\sigma$, the boudary approximates the spherical boundary.

\TODO{More on SVM parameters, \cite{cherkassky2007learning}, section 9.8, page 446}

\TODO{Analysis of C/$\nu$-parameter, which sets a lower bound for the fraction of support vectors and an uppor bound for samples that will be outliers. \cite{flach2012machine} section 9.8 and \cite{scholkopf2002learning} and other Sch\"olkopf material.}

@@
